---
title: "R Notebook"
output: html_notebook
---

```{r}
### Read training data
#! Perhaps you need to set the working directory!?
#setwd("/home/pbac/g/course02417/2025/assignment1")
D <- read.csv("DST_BIL54.csv")
str(D)

# See the help
?strftime
D$time <- as.POSIXct(paste0(D$time,"-01"), "%Y-%m-%d", tz="UTC")
D$time
class(D$time)

## Year to month for each of them
D$year <- 1900 + as.POSIXlt(D$time)$year + as.POSIXlt(D$time)$mon / 12

## Make the output variable a floating point (i.e.\ decimal number)
D$total <- as.numeric(D$total) / 1E6

## Divide intro train and test set
teststart <- as.POSIXct("2024-01-01", tz="UTC")
Dtrain <- D[D$time < teststart, ]
Dtest <- D[D$time >= teststart, ]
```

```{r}
Dtrain
```
## 1. Plot data

1.1
```{r}
dates <- as.Date(Dtrain$time)

# Create x 
Dtrain$x <- as.numeric(format(Dtrain$time, "%Y")) + 
            (as.numeric(format(Dtrain$time, "%m")) - 1) / 12
```

1.2
```{r}
# Plot training data vs x

variable_of_interest = Dtrain$total
plot(Dtrain$x, variable_of_interest, type = "l", xlab = "Time Variable X", 
     ylab = "Total", main = "Total vs Time Variable X")

```
```{r}
# Basic statistics
cat("Start value (2018-01):", Dtrain$total[1], "\n") 
cat("End value (2023-12):", tail(Dtrain$total, 1), "\n")
cat("Total growth:", tail(Dtrain$total, 1) - Dtrain$total[1], "\n")
cat("Monthly observations:", nrow(Dtrain), "\n")
cat("Average monthly growth:", mean(diff(Dtrain$total)), "\n")
cat("Annual growth rate:", mean(diff(Dtrain$total)) * 12, "\n")

# Linear trend fit (R² for predictability)
fit <- lm(total ~ x, data = Dtrain)
cat("Linear model R²:", round(summary(fit)$r.squared, 3), "\n")

# Print results
print(fit)
```

The training data plot shows a clear upward trend in total registered motor vehicles in Denmark from 2018 to approximately 2022 with slight downward hills in beginning of 2022 and also in the end of 2022 leading up to the start of 2023. 


## 2. Linear trend model

```{r}
head(Dtrain[, c("x", "total")], 3)
```

## 3. OLS - global linear trend model

# 3.1. Estimate the parameters θ1 and θ2 using the training set (call it the Ordinary Least Squares (OLS) estimates). Describe how you calculated the estimates.

```{r}
y <- Dtrain$total
X <- cbind(1, Dtrain$x)

thetahat <- solve(t(X) %*% X) %*% t(X) %*% y

thetahat

```

# 3.2

```{r}

# Step 1: Residuals
residuals <- y - X %*% thetahat

# Step 2: Estimate sigma^2 (residual variance)
n <- nrow(X)           # 72 observations
p <- ncol(X)           # 2 parameters
sigma2_hat <- sum(residuals^2) / (n - p)  # unbiased estimator

# Step 3: Variance-covariance matrix of thetahat
V_theta <- sigma2_hat * solve(t(X) %*% X)

# Step 4: Standard errors = sqrt(diag(V_theta))
se_theta1 <- sqrt(V_theta[1,1])  # σ̂_θ₁
se_theta2 <- sqrt(V_theta[2,2])  # σ̂_θ₂

se_theta1

se_theta2
```

```{r}
fitted_mean <- X %*% thetahat  # μ̂(x) for all training points

plot(Dtrain$x, Dtrain$total, 
     pch = 16, col = "steelblue", cex = 0.8,
     xlab = "Time Variable x (fractional years)",
     ylab = "Total Registered Vehicles",
     main = "Observations and Estimated Mean")

lines(Dtrain$x, fitted_mean, col = "red", lwd = 2)

# Legend
legend("topleft", 
       legend = c("Observations y", "Estimated Mean)"), 
       pch = c(16, NA), 
       lty = c(NA, 1), 
       col = c("steelblue", "red"), 
       lwd = c(NA, 2))

```

# 3.3 Make a forecast for the test set, hence the following 12 months - i.e., compute predicted values
with corresponding prediction intervals for 2024-Jan to 2024-Dec. Present these values in a table.

```{r}
# Test set: 2024-Jan to Dec

# Create x 
Dtest$x <- as.numeric(format(Dtest$time, "%Y")) + 
            (as.numeric(format(Dtest$time, "%m")) - 1) / 12

X_test <- cbind(1, Dtest$x)

# Predictions + 95% prediction intervals
yhat_test <- X_test %*% thetahat

se_pred_test <- sqrt(sigma2_hat + diag(X_test %*% V_theta %*% t(X_test)))
t_crit <- qt(0.975, df = nrow(X) - ncol(X))  # 95% critical t-value
pred_int <- cbind(
  lower = yhat_test - t_crit * se_pred_test,
  upper = yhat_test + t_crit * se_pred_test
)

# Forecast table using Dtest months
forecast_table <- data.frame(
  Month = format(Dtest$time, "%b %Y"),
  x = round(Dtest$x, 3),
  Actual = round(Dtest$total, 3),
  Forecast = round(yhat_test, 3),
  `Lower 95%` = round(pred_int[,1], 3),
  `Upper 95%` = round(pred_int[,2], 3)
)
print(forecast_table, row.names = FALSE)
```


# 3.4 Plot the fitted model together with the training data and the forecasted values (also plot the
prediction intervals of the forecasted values).

```{r}
# TOD
```

# 3.6 Investigate the residuals of the model

```{r}
# Todo
```


